\bf Model Visualization Techniques for a Social Network Model

### 2017-09-15 paper.Rnw
making progress

We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, $M1$, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, $M2$ and $M3$, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The $M2$ model contains an actor-level covariate parameter, and the $M3$ model contains an additional strutural effect in the objective function.
\begin{align*}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4}, 
\end{align*}

@@

For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data. Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}. In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:

@@

In $M4$, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In $M5$,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. This parameter attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties. By exploring these models more visually, we hope to understand these effects and the model fitting process better.

@@

% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model. On the right, the second wave of observed data. In the middle, a summary network from the first model fit to the data. This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model $M1$.',results='hide', fig.height=3>>=
library(netvizinf)
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- s50a[20:35,1]
drink2 <- s50a[20:35,2]
drink3 <- s50a[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M1eff,
                       parms = as.numeric(M1parms), 
                       N = 1000)
M2sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M2eff,
                       parms = as.numeric(M2parms), 
                       N = 1000)
M3sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M3eff,
                       parms = as.numeric(M3parms), 
                       N = 1000)
M1simsdf <- sims_to_df(M1sims1000)
M2simsdf <- sims_to_df(M2sims1000)
M3simsdf <- sims_to_df(M3sims1000)
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure. Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure. We could talk about expected values of parameters, but the parameters can be hard to interpret. Furthermore, there is no way to talk about the expected value of an observation from a statistical network model. How then, can we arrive at an ``average" network? We propose to answer this question through visualization. For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the average network is much more similar to the first wave than to the second wave. However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that the simple model, $M1$, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

@@

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model $M1$.}
\end{figure}

Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}. A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p. 4369]{buja:2009}. It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect. In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of mock data, simulated from a hypothesized model. If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit. \st{An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences. They pose questions to experiment participants such as, ``Which plot is most different from the others?" for the first example and, ``In which plot is the blue group furthest to the right?" for the distributional differences. The data visualizations examined in \citet{hofmann2012graphical} are less complex than a node-link network diagram. What questions should we ask for network data visualizations? Asking participants to identify the most different plot may be difficult, as we show} in Figure~\ref{fig:lineup}. In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data. What makes these plots ``different"? It seems possible to argue for any one of the six plots in Firgure~\ref{fig:lineup} as most different. So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure. The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data. We have found in an experiment where all the node-link diagrams shown are simulated, one from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out to people the most. We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.}

@@

To our small friendship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each. We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}. We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties. It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.

@@

<<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side. There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=6>>=
#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(drink[20:35,])
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
data(M1ests_bigfriends)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
# M1sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M1eff,
#                        parms = as.numeric(M1parms), 
#                        N = 1000)
# M2sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M2eff,
#                        parms = as.numeric(M2parms), 
#                        N = 1000)
# M3sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M3eff,
#                        parms = as.numeric(M3parms), 
#                        N = 1000)
# M1simsdf <- sims_to_df(M1sims1000)
# M2simsdf <- sims_to_df(M2sims1000)
# M3simsdf <- sims_to_df(M3sims1000)
M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))

# M1avg$from <- paste0("V", M1avg$from)
# M1avg$to <- paste0("V", M1avg$to)
# M1avg2$from <- paste0("V", M1avg2$from)
# M1avg2$to <- paste0("V", M1avg2$to)
# M1avgW2$from <- paste0("V", M1avgW2$from)
# M1avgW2$to <- paste0("V", M1avgW2$to)
# M2avg$from <- paste0("V", M2avg$from)
# M2avg$to <- paste0("V", M2avg$to)
# M2avgW2$from <- paste0("V", M2avgW2$from)
# M2avgW2$to <- paste0("V", M2avgW2$to)
# M2avg2$from <- paste0("V", M2avg2$from)
# M2avg2$to <- paste0("V", M2avg2$to)
# M3avg$from <- paste0("V", M3avg$from)
# M3avg$to <- paste0("V", M3avg$to)
# M3avgW2$from <- paste0("V", M3avgW2$from)
# M3avgW2$to <- paste0("V", M3avgW2$to)
# M3avg2$from <- paste0("V", M3avg2$from)
# M3avg2$to <- paste0("V", M3avg2$to)
combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
ggplot(data = combinedavgs) + 
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) + 
  ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
  facet_wrap(~cat, nrow = 2)

@@

<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id. On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
# reinstall new dplyr
numordersf <- data.frame(fd2.w1) %>% 
  mutate(from = rownames(fd2.w1)) %>% 
  gather(key = to ,value = value, -from)
pcaordersf <- numordersf

@@

<<adjmatani, fig.cap='The adjacency matrix visualization animation for one series of microsteps. At the beginning of the animation, there are two connected components. By the end, the components appear to be more spread out.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}

>>>>

We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align*}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4}, 
\end{align*}

@@

For comparison, we also fit models M1, M2, and M3 to the senate collaboration data. Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}. In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. This parameter attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties. By exploring these models more visually, we hope to understand these effects and the model fitting process better.

@@

% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model. On the right, the second wave of observed data. In the middle, a summary network from the first model fit to the data. This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=3>>=
library(netvizinf)
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- s50a[20:35,1]
drink2 <- s50a[20:35,2]
drink3 <- s50a[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M1eff,
                       parms = as.numeric(M1parms), 
                       N = 1000)
M2sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M2eff,
                       parms = as.numeric(M2parms), 
                       N = 1000)
M3sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M3eff,
                       parms = as.numeric(M3parms), 
                       N = 1000)
M1simsdf <- sims_to_df(M1sims1000)
M2simsdf <- sims_to_df(M2sims1000)
M3simsdf <- sims_to_df(M3sims1000)
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure. Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure. We could talk about expected values of parameters, but the parameters can be hard to interpret. Furthermore, there is no way to talk about the expected value of an observation from a statistical network model. How then, can we arrive at an ``average" network? We propose to answer this question through visualization. For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the average network is much more similar to the first wave than to the second wave. However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

@@

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model M1.}
\end{figure}

Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}. A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p. 4369]{buja:2009}. It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect. In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of mock data, simulated from a hypothesized model. If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit. \st{An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences. They pose questions to experiment participants such as, ``Which plot is most different from the others?" for the first example and, ``In which plot is the blue group furthest to the right?" for the distributional differences. The data visualizations examined in \citet{hofmann2012graphical} are less complex than a node-link network diagram. What questions should we ask for network data visualizations? Asking participants to identify the most different plot may be difficult, as we show} in Figure~\ref{fig:lineup}. In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data. What makes these plots ``different"? It seems possible to argue for any one of the six plots in Firgure~\ref{fig:lineup} as most different. So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure. The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data. We have found in an experiment where all the node-link diagrams shown are simulated, one from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out to people the most. We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.}

@@

To our small friendship network, we fit three models, M1, M2, and M3, using \texttt{RSiena} 1,000 times each. We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}. We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. \st{When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls. The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently. The outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases. This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends. Having many friends who do not reciprocate is discouraged by M2.} In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties. It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.

@@

<<averagenet, results='hide', out.width='99%',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above. There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=6>>=
#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(drink[20:35,])
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
data(M1ests_bigfriends)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
# M1sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M1eff,
#                        parms = as.numeric(M1parms), 
#                        N = 1000)
# M2sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M2eff,
#                        parms = as.numeric(M2parms), 
#                        N = 1000)
# M3sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M3eff,
#                        parms = as.numeric(M3parms), 
#                        N = 1000)
# M1simsdf <- sims_to_df(M1sims1000)
# M2simsdf <- sims_to_df(M2sims1000)
# M3simsdf <- sims_to_df(M3sims1000)
M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))

# M1avg$from <- paste0("V", M1avg$from)
# M1avg$to <- paste0("V", M1avg$to)
# M1avg2$from <- paste0("V", M1avg2$from)
# M1avg2$to <- paste0("V", M1avg2$to)
# M1avgW2$from <- paste0("V", M1avgW2$from)
# M1avgW2$to <- paste0("V", M1avgW2$to)
# M2avg$from <- paste0("V", M2avg$from)
# M2avg$to <- paste0("V", M2avg$to)
# M2avgW2$from <- paste0("V", M2avgW2$from)
# M2avgW2$to <- paste0("V", M2avgW2$to)
# M2avg2$from <- paste0("V", M2avg2$from)
# M2avg2$to <- paste0("V", M2avg2$to)
# M3avg$from <- paste0("V", M3avg$from)
# M3avg$to <- paste0("V", M3avg$to)
# M3avgW2$from <- paste0("V", M3avgW2$from)
# M3avgW2$to <- paste0("V", M3avgW2$to)
# M3avg2$from <- paste0("V", M3avg2$from)
# M3avg2$to <- paste0("V", M3avg2$to)
combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
combinedavgs$cat <- as.factor(combinedavgs$cat)
combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))

ggplot(data = combinedavgs) + 
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) + 
  ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
  facet_wrap(~cat)

@@

<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id. On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
# reinstall new dplyr
numordersf <- data.frame(fd2.w1) %>% 
  mutate(from = rownames(fd2.w1)) %>% 
  gather(key = to ,value = value, -from)
pcaordersf <- numordersf

@@

<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps. At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner. By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}
